---
title: "Assignment 1"
author: "Greg Strohl"
date: "10/20/2019"
output:
  word_document: 
  keep_md: yes
html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
```



#### Explain your code where appropriate.

Reference: [UCI Machine Learning Repository: Wine Dataset](https://archive.ics.uci.edu/ml/datasets/Wine)

The wine data set consists of chemical measurements on 13 constituents found in each of the three types of wines (cultivars) grown in the same region of Italy.

1. Compute the Alcohol means by `Cultivar`. Briefly discuss the mean differences among the cultivars.
```{r message=TRUE, warning=TRUE, paged.print=TRUE}
Cultivar<-as.factor(Wine[,"Cultivar"])
Alcohol<-Wine[,"Alcohol"]
# Split `alcohol` by `cultivar` to form a list with each element containing the data for a group.
# Look at the function `sapply`.
# Put your R code here.
AlcbyCult<-cbind(Cultivar,Alcohol)
sapply(split(Alcohol,Cultivar),mean)
alcoholList<-split(Alcohol,Cultivar)
sapply(alcoholList,mean)
summary(Wine)
plot(Alcohol~Cultivar)
```

I combined the alcohol and cultivar into a matrix called $AlcByCult$ using the 'cbind' function.  From there, I used the 'sapply' function to calculate the mean alcohol per cultivar.

The mean alcohol for cultivar 1 was 13.74475. 
The mean alcohol for cultivar 2 was 12.27873. 
The mean alcohol for cultivar 3 was 13.15375.

2. Compute the number of observations in each `cultivar`.

```{r}
library(dplyr)
dim(Wine)
arrange(Wine)
Wine %>%
  group_by(Cultivar) %>%
  summarize(n())
sapply(alcoholList,length)
```
Using the wine data set, I used the dplyr function to group and summarize the data by Cultivar.
There were 59 observations for cultivar 1.
There were 71 observations for cultivar 2.
There were 48 observations for cultivar 3.

3.  Create a function to perform a one-way analysis of variance. The input argument `z` should be a list consisting of (possibly) named components, one for each group. The output should be a named list containing components for the between SS ($SS_B$), the within SS ($SS_W$), the between degrees of freedom, and the within degrees of freedom.  
Note: $SS_B = \sum_i n_i (\bar{y}_i - \bar{y})^2$ and $SS_W = \sum_i (n_i - 1) s_i^2$ where $n_i$ is the sample size of group $i$, $\bar{y}_i$ is the mean of group $i$, and  $s_i^2$ is the variance of group $i$. These group statistics can easily be computed using `sapply`. For the grand mean, $\bar{y}$, think about using `unlist` on `z`. Let $g$ be the number of groups and $n = \sum_i n_i$ be the total sample size, which can also be computed by `unlist`ing `z`. $n$ and $g$ are needed to compute the between and within degrees of freedom.  
Note: The code should be general for any $g$ and $n_i$.


```{r}
# Look at sapply for summarizing over the elements of a list.
oneway <- function(z){
  # Put your R code here.
  summary(Wine)
  n <- length(unlist(z,recursive = TRUE))
  n_i <-sapply(z,length)
  s_i <- sapply(z,var)
  g <- length(z)
  y_bar <-mean(unlist(z,recursive = TRUE))
  y<- sapply(z,mean)
  ssb<- n_i*(sapply(z,mean)-y_bar)^2
  ssw<- sum(n_i-1)*s_i^2
  return(list(ssb=sum(ssb),ssw=sum(ssw),n=n,g=g))  
}
x<-oneway(alcoholList)
x
df1 = x[[4]]-1
df2 = x[[3]]-x[[4]]
p = pf(x[[2]],df1,df2)
df1
df2
p
```

4.  Create a function to summarize the output in a one-way ANOVA table, including the F test and $p$-value. The input argument is the output named list in the previous question. The output should be one-way ANOVA table.  
Note: For computing the $p$-value look at the R function `pf`.
```{r, message=FALSE, warning=FALSE}
# For your output, mimic the tabular output of the builtin `summary` function applied to the output of the builtin `aov` function.
# Look at the function `printCoefmat` to form a table.
oneway.table <- function(x){
  # Put your R code here.
  df1 = x[[4]]-1
  df2 = x[[3]]-x[[4]]
  p = pf(x[[1]],df1,df2)
  ss = sum(x[[1]]^2)
}
```

5. Your functions should be illustrated with the `wine` data set. The data consists of 178 samples measuring alcohol (the outcome variable) divided among three (3) cultivars (the input variable).


```{r, message=FALSE, warning=FALSE}
# Split `alcohol` by `cultivar` to call `oneway`.
# Put your R code here.
attach(Wine)
data(Wine)
str(Wine)
# Summary of the analysis
wine.aov <- aov(Cultivar~Alcohol, data = Wine)
summary(wine.aov)
model1<- aov(Cultivar ~ Alcohol)
par(mfrow=c(2,2))
plot(wine.aov, 2)
boxplot(Wine$Alcohol ~ Cultivar, 
        vertical = TRUE, 
        main="AlcByCult",
        col = "blue")
```

